---
title: "ND GAIN Data Exploration"
author: "Connor"
date: "6/13/2023"
output: html_document
---
```{r}

```

Load Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)# allows us to cut out long file paths (ex. "users/connor/dowloads/etc")
library(tidyverse) # allows us to do lots of fundamental data science tasks (ex. read_csv)
library(ggcorrplot) # allows us to make correlation plots
library(plotly) # allows us to make ggplot objects interactive
library(ggpubr) # stat_cor for correlation coefficients
library(janitor)
library(sf) #working with spatial data 
library(leaflet) #creates interactive maps 
library (rnaturalearth) # accessing natural earth data
library(readr)
library(rnaturalearthdata)
library(ggrepel) # extend ggplot 2 package for overlapping text labels 
library(countrycode)
library(rgdal)
```
Read Data
[ND-GAIN Country Index](https://gain.nd.edu/our-work/country-index/)
The ND-GAIN Country Index summarizes a country's vulnerability to climate change and other global challenges in combination with its readiness to improve resilience. It aims to help governments, businesses and communities better prioritize investments for a more efficient response to the immediate global challenges ahead.

```{r} 
nd_gain_data <- read_csv(here("SIDS_Data_Platform_ND-GAIN/data/nd-gain-joined.csv"))

```


Round value column for two numbers after the decimal

```{r}
nd_gain_data$value <- round(nd_gain_data$value, digits = 2)
```


**Challenge**

The dataset is slightly flawed because if a value is not measured one year, it simply gets carried over from the previous year

So simply think of filtering for the year 2020 as using the most recent year of available data (warning: this can lead to 20 year old data being visualized)


```{r}
nd_gain_most_recent <- nd_gain_data %>% 
  filter(year == 2020)
```

Shift dataframe so that each value in the indicator column becomes it's own column 
 - This gives us 190 rows (one country per row) with 48 indicators

```{r}
nd_gain_most_recent_wider <- nd_gain_most_recent %>% 
  pivot_wider(names_from = indicator,
              values_from = value)
```

Now our data is ready to visualize!

Here are a few interesting plots created by our SPICE 2023 Summer Students funded by the Pacific Intelligence and Innovation Initiative (P3I)


1. Histogram

This code chunk below creates a histogram that shows the distribution of dependency on external resource for health services across 190 countries colored by whether the country is Small Island Developing State (SIDS) or not

```{r}
ggplot(data = nd_gain_most_recent_wider, aes(x = dependency_on_external_resource_for_health_services,
                                     fill = sids_context
                                     )) +
  geom_histogram(bins = 65, color = "black") +
  theme_minimal()

```{r}
penguins_lter <- penguins_lter %>% 
clean_names()
```


2. Scatter Plots

Create a scatter plot with education on the x axis and the human development index on the y axis

```{r}
ggplot(data = nd_gain_most_recent_wider, aes(x = education,
                                     y = human_development_index)) +
  geom_point() +
  labs(title = "Education and Human Development Index, ND-GAIN Most Recent Data",
       x = "education",
       y = "human development index") +
  theme_minimal()
```


```{r}
ggplot(data = penguins_lter_torgerson, aes(x = flipper_length_mm,
                                     y = body_mass_g,
                                 color = sex,
                                 shape = stage)) +
  geom_point() +
  labs(title = "Body Mass of Penguins in Torgersen",
       x = "flipper length",
       y = "Body Mass (g)") +
  theme_minimal()
```
```{r}

```

```
Now add a trend line with geom_smooth() and a p value and correlation coefficient with stat_cor()
```{r}
ggplot(data=nd_gain_most_recent_wider, aes(x=ecological_footprint,
                                            y=human_development_index))+
geom_point()+
  labs(title = "Ecological Footprint and Human Development Index, ND-GAIN Most Recent Data",
       x = "ecological_footprint",
       y = "human_development_index") +
  theme_minimal()
```


```{r}
ggplot(data = nd_gain_most_recent_wider, aes(x = education,
                                     y = human_development_index)) +
  geom_point() +
  geom_smooth() +
  stat_cor(method = "pearson", output.type = "text", label.sep='\n') +
  labs(title = "Education and Human Development Index, ND-GAIN Most Recent Data",
       x = "education",
       y = "human development index") +
  theme_minimal()
```
```{r}
ggplot(data = nd_gain_most_recent_wider, aes(x = ecological_footprint,
                                     y = projected_population_change)) +
  geom_point() +
  geom_smooth() +
  stat_cor(method = "pearson", output.type = "text", label.sep='\n') +
  labs(title = "Ecological Footprint and Projected Population Change, ND-GAIN Most Recent Data",
       x = "ecological_footprint",
       y = "projected_population_change") +
  theme_minimal()
  
```

Now split the plot into two for Small Island Developing States (SIDS) and non-SIDS 
 - Warning -> Be cautious of using trend lines, p-values, and correlation coefficients with <30 data points (in this case countries)

```{r}
ggplot(data = nd_gain_most_recent_wider, aes(x = education,
                                     y = human_development_index)) +
  geom_point() +
  geom_smooth() +
  stat_cor(method = "pearson", output.type = "text", label.sep='\n') +
  facet_wrap(~sids_context) +
  labs(title = "Education and Human Development Index, ND-GAIN Most Recent Data",
       x = "education",
       y = "human development index") +
  theme_minimal()
```

```{r}
ggplot(data = nd_gain_most_recent_wider, aes(x = ecological_footprint,
                                     y = projected_population_change)) +
  geom_point() +
  geom_smooth() +
  stat_cor(method = "pearson", output.type = "text", label.sep='\n') +
  facet_wrap(~sids_context) +
  labs(title = "Ecological Footprint and Projected Population Change, ND-GAIN Most Recent Data",
       x = "ecological footprint",
       y = "projected population change") +
  theme_minimal()
  
```


Now add color indicating rural population
```

```

```{r}

Lastly, make the plot interactive, adding label = Name so we can see which point is which country by hovering, and by naming it and inputting it into ggplotly()


```{r}
education_hdi_gdp_plot <- ggplot(data = nd_gain_most_recent_wider, aes(x = education,
                                     y = human_development_index,
                                     color = rural_population,
                                     label = Name)) +
  geom_point() +
  geom_smooth() +
  stat_cor(method = "pearson", output.type = "text", label.sep='\n', label.x = 20) +
  facet_wrap(~sids_context) +
  labs(title = "Education and Human Development Index, ND-GAIN Most Recent Data",
       x = "education",
       y = "human development index") +
  theme_minimal()

ggplotly(education_hdi_gdp_plot)
```


3. Interactive Scatter Plot without Trend Lines
- Sometimes, we don't intend to show a trend with a scatter plot


Use a scatter plot to display projected change of sea level rise and engagement in international environmental conventions, color by SIDS or non-SIDS


```{r}
engagement_impacts_plot <- ggplot(data = nd_gain_most_recent_wider, aes(x = engagement_in_international_environmental_conventions, 
                                             y = projected_change_of_sea_level_rise_impacts, 
                                             color = sids_context,
                                             label = Name)) +
  geom_point() +
  labs(title = "Projected Change of Sea level Rise Impacts / Engagement in International Environmental Conventions",
       subtitle = "ND-GAIN Data",
       x = "engagement in international environmental conventions",
       y = "projected change of sea level rise impacts") +
  theme_minimal()

ggplotly(engagement_impacts_plot)
```
```{r}
engagement_impacts_plot <- ggplot(data = nd_gain_most_recent_wider, aes(x = ecological_footprint, 
                                             y = projected_population_change, 
                                             color = sids_context,
                                             label = Name)) +
  geom_point() +
  labs(title = "Ecological Footprint and Projected Population Change",
       subtitle = "ND-GAIN Data",
       x = "ecological footprint",
       y = "projected population change") +
  theme_minimal()

ggplotly(engagement_impacts_plot)


```
```{r}
engagement_impacts_plot <- ggplot(data = nd_gain_most_recent_wider, aes(x = urban_concentration, 
                                             y = projected_change_of_warm_periods, 
                                             color = sids_context,
                                             label = Name)) +
  geom_point() +
  labs(title = "Projected Change of Warm Periods and Urban Concentration, SIDS",
       subtitle = "ND-GAIN Data",
       x = "urban concentration",
       y = "projected change of warm periods") +
  theme_minimal()

ggplotly(engagement_impacts_plot)

```



4. Subset Data and Create a Bar Chart

Subset Data for only SIDS

```{r}
nd_gain_most_recent_wider_sids <- nd_gain_most_recent_wider %>% 
  filter(sids_context == "sids")
```

```{r}
nd_gain_most_recent_wider_non_sids <- nd_gain_most_recent_wider %>%
  filter(sids_context == "non-sids")
```
```{r}
penguins_lter_Island <- penguins_lter %>%
  filter(Island == "Dream")
```

Create Bar Chart showing Food Import Dependency in SIDS 

```{r}
ggplot(data = nd_gain_most_recent_wider_non_sids, aes(x = food_import_dependency, 
                                                  y = reorder(Name,food_import_dependency))) + 
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(title = "Food Import Dependency, SIDS",
       x = "proportion of cereal consumption obtained from imports",
       y = "")+
  theme_minimal() 
  
```
```{r}
ggplot(data = nd_gain_most_recent_wider_sids, aes(x = projected_change_of_warm_periods, 
                                                  y = reorder(Name,projected_change_of_warm_periods))) + 
  geom_bar(stat = "identity", fill = "purple", color = "black") +
  labs(title = "Projected Change of Warm Periods, SIDS",
       x = "urban concentration",
       y = "projected change of warm periods")+
  theme_minimal() 
```{r}
ggplot(data = penguins_lter, aes(x = island, 
                                 y = 
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(title = "Date of Eggs and Region",
       x = "Date Egg",
       y = "Islands")+
  theme_minimal() 
```

Leaflet Map - downloading rnaturalearth
```{r}
countries <- ne_countries(scale = "medium", returnclass = "sf")
```

Plotting Countries 
```{r}
plot(countries)
```



```{r}
countries_df <- as.data.frame(countries)
```


```{r}
plot(countries_df$sovereignt)
```

```{r}

y_lower_limit <- 0
y_upper_limit <- 100

```

```{r}

plot(countries, ylim = c(y_lower_limit, y_upper_limit))


```



plot(countries, ylim = c(y_lower_limit, y_upper_limit))





Creating a Map - Choropleth

 read data from the CSV file 
```{r}

data <- read_csv("C:\\Users\\chosu\\OneDrive\\Documents\\Data-4-SDG\\SIDS_Data_Platform_ND-GAIN\\data\\data.csv")
                         
```
 get world country polygons 
```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
```

merge country polygons with the data 
```{r}
merged_data <- merge(world, data, by.x = "name", by.y = "country", all.x = TRUE )
```

create a leaflet map 
```{r}
map <- leaflet(data = merged_data) %>%
  addTiles()  
```
adding default openstreetmap tiles as the base layer^

add markers with popups for each location and display multiple variables in the popup 
```{r}
map <- map %>%
  addPolygons(
    fillColor = "blue",
    fillOpacity = 0.5,
    color = "white",
    weight = 1,
    popup = ~paste("Country ", name, "<br>",
                   "Sea Level Impact: ", projected_change_of_sea_level_rise_impacts, "<br>",
                   "Disaster Preparedness: ", disaster_preparedness),
    label = ~as.character(iso_a3),
    labelOptions = labelOptions(noHide = TRUE)
  )
```

```{r}
map
```

Retry 

Get world data: 
```{r}
world <- ne_countries(scale="small", returnclass = "sf")
```

plot an empty world map 
```{r}
world %>%
  ggplot()+
  geom_sf()
```
Change map projection
 #"+proj=robin", "+proj=moll", "+proj=aeqd", "+proj=goode"
```{r}
world %>% 
  st_transform(crs = "+proj=robin") %>%
  ggplot() +
  geom_sf() +
  theme_minimal()
```

remove the graticule 
```{r}
world %>%
  st_transform(crs = "+proj=wintri") %>%
  ggplot() +
  geom_sf() +
  coord_sf(datum = NA) +
  theme_minimal()
```

prepare data 
```{r}
data_raw <- read_csv ("C:\\Users\\chosu\\OneDrive\\Documents\\Data-4-SDG\\SIDS_Data_Platform_ND-GAIN\\data\\data.csv")
```
tidy dataset in long format 
```{r}
data <- data_raw %>%
  select(projected_change_of_sea_level_rise_impacts, disaster_preparedness, country) %>%
  mutate(visited = TRUE )
```

add iso3 country code
```{r}
data_with_iso <- data %>%
  mutate(Iso3 = countrycode::countrycode(
    sourcevar = country,
    origin = "country_code",
    destination = "iso3c"
))
```
```{r}
data_with_iso <- data %>%
  mutate(Iso3 = countrycode(country, "country.name", "iso3c"))
```


Different Method by Connor 
```{r}
library(leaflet)
```
```{r}
# Download the shapefile. (note that I store it in a folder called DATA. You have to change that if needed.)
download.file("http://thematicmapping.org/downloads/TM_WORLD_BORDERS_SIMPL-0.3.zip" , destfile="DATA/world_shape_file.zip")
# You now have it in your current working directory, have a look!

# Unzip this file. You can do it with R (as below), or clicking on the object you downloaded.
system("unzip DATA/world_shape_file.zip")
#  -- > You now have 4 files. One of these files is a .shp file! (TM_WORLD_BORDERS_SIMPL-0.3.shp)
```
```{r}
# Read this shape file with the rgdal library. 

world_spdf <- readOGR(here("SIDS_Data_Platform_ND-GAIN/TM_WORLD_BORDERS_SIMPL-0.3.shp"))


# Clean the data object
library(dplyr)
world_spdf@data$POP2005[ which(world_spdf@data$POP2005 == 0)] = NA
world_spdf@data$POP2005 <- as.numeric(as.character(world_spdf@data$POP2005)) / 1000000 %>% round(2)
# Create a color palette for the map:
mypalette <- colorNumeric( palette="viridis", domain=world_spdf@data$POP2005, na.color="transparent")
mypalette(c(45,43))

# Basic choropleth with leaflet?
m <- leaflet(world_spdf) %>% 
  addTiles()  %>% 
  setView( lat=10, lng=0 , zoom=2) %>%
  addPolygons( fillColor = ~mypalette(POP2005), stroke=FALSE )

m
```
```{r}
dataframe_world <- as.data.frame(world_spdf@data)
```

```{r}
data_with_iso <- data_with_iso %>%
  rename(ISO3 = Iso3)

```
```{r}
data_joined <- full_join(data_with_iso, dataframe_world, by = "ISO3")
```

